---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r srr-tags, eval = FALSE, echo = FALSE}
# TODO all
#' @srrstats {G1.1} *Statistical Software should document whether the algorithm(s) it implements are:* - *The first implementation of a novel algorithm*; or - *The first implementation within **R** of an algorithm which has previously been implemented in other languages or contexts*; or - *An improvement on other implementations of similar algorithms in **R***.
#' @srrstats {G1.2} *Statistical Software should include a* Life Cycle Statement *describing current and anticipated future states of development.* 
#' @srrstats {G1.3} *All statistical terminology should be clarified and unambiguously defined.*
#' @srrstats {G1.4} *Software should use [`roxygen2`](https://roxygen2.r-lib.org/) to document all functions.*
#' @srrstats {G1.4a} *All internal (non-exported) functions should also be documented in standard [`roxygen2`](https://roxygen2.r-lib.org/) format, along with a final `@noRd` tag to suppress automatic generation of `.Rd` files.*


#' @srrstats {G2.0} *Implement assertions on lengths of inputs, particularly through asserting that inputs expected to be single- or multi-valued are indeed so.*
#' @srrstats {G2.0a} *Provide explicit secondary documentation of any expectations on lengths of inputs*
#' @srrstats {G2.1} *Implement assertions on types of inputs (see the initial point on nomenclature above).*
#' @srrstats {G2.1a} *Provide explicit secondary documentation of expectations on data types of all vector inputs.*
#' @srrstats {G2.2} *Appropriately prohibit or restrict submission of multivariate input to parameters expected to be univariate.*
#' @srrstats {G2.3} *For univariate character input:*
#' @srrstats {G2.3a} *Use `match.arg()` or equivalent where applicable to only permit expected values.*
#' @srrstats {G2.3b} *Either: use `tolower()` or equivalent to ensure input of character parameters is not case dependent; or explicitly document that parameters are strictly case-sensitive.*
#' @srrstats {G2.6} *Software which accepts one-dimensional input should ensure values are appropriately pre-processed regardless of class structures.*
#' @srrstats {G2.7} *Software should accept as input as many of the above standard tabular forms as possible, including extension to domain-specific forms.*
#' @srrstats {G3.0} *Statistical software should never compare floating point numbers for equality. All numeric equality comparisons should either ensure that they are made between integers, or use appropriate tolerances for approximate equality.*
#' @srrstats {G5.0} *Where applicable or practicable, tests should use standard data sets with known properties (for example, the [NIST Standard Reference Datasets](https://www.itl.nist.gov/div898/strd/), or data sets provided by other widely-used R packages).* Grunfeld data example?
#' @srrstats {BS1.2a} *The main package `README`, either as textual description or example code*
#' @srrstats {BS1.2c} *Function-level documentation, preferably with code included in examples*
# TODO comparing strings?
# Placeholder for the potential ropensci software review tags
```

# dynamite

<!-- badges: start -->
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![R-CMD-check](https://github.com/santikka/dynamite/workflows/R-CMD-check/badge.svg)](https://github.com/santikka/dynamite/actions)
[![Codecov test coverage](https://codecov.io/gh/santikka/dynamite/branch/main/graph/badge.svg)](https://app.codecov.io/gh/santikka/dynamite?branch=main)
<!-- badges: end -->

The `dynamite` package provides easy-to-use interface for Bayesian inference of complex panel data. The main features distinguishing the package and the underlying methodology from many other approaches are 

* Support for both time-varying and time-invariant effects
* Joint modelling of multiple measurements per individual (multiple channels)
* Support for non-gaussian observations.
* Realistic counterfactual predictions which take account the dynamic structure of the model
* Clear quantification of parameter and predictive uncertainty due to Bayesian approach
* User-friendly and efficient R interface with state-of-the-art estimation via Stan.


The `dynamite` package is developed with the support of Academy of Finland grant 331817. 

## Installation

You can install the development version of dynamite from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("santikka/dynamite")
```

## Example

TODO
