---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "50%"
)
```

```{r srr-tags, eval = FALSE, echo = FALSE}
#' @srrstats {G1.1,  G1.2}
```

# dynamite

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/santikka/dynamite/workflows/R-CMD-check/badge.svg)](https://github.com/santikka/dynamite/actions)
[![Codecov test coverage](https://codecov.io/gh/santikka/dynamite/branch/main/graph/badge.svg)](https://app.codecov.io/gh/santikka/dynamite?branch=main)
<!-- badges: end -->

The `dynamite` package provides easy-to-use interface for Bayesian inference of complex panel data. The main features distinguishing the package and the underlying methodology from many other approaches are:

* Support for both time-varying and time-invariant effects.
* Joint modeling of multiple measurements per individual (multiple channels).
* Support for non-gaussian observations.
* Realistic counterfactual predictions which take into account the dynamic structure of the model.
* Clear quantification of parameter and predictive uncertainty due to a Bayesian approach.
* User-friendly and efficient R interface with state-of-the-art estimation via Stan.

The `dynamite` package is developed with the support of Academy of Finland grant 331817. 

## Installation

You can install the development version of dynamite from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("santikka/dynamite")
```

## Example

A single-channel model with time-invariant effect of `z`, time-varying effect 
of `x`, lagged value of the response variable `y` and a group-specific random 
intercepts:

```{r, echo = FALSE}
library(dynamite)
```

```{r, eval = FALSE}
set.seed(1)
library(dynamite)
gaussian_example_fit <- dynamite(
  obs(y ~ -1 + z + varying(~ x + lag(y)), family = "gaussian",
      random_intercept = TRUE) + splines(df = 20),
  data = gaussian_example, time = "time", group = "id",
  iter = 2000, warmup = 1000, thin = 5,
  chains = 2, cores = 2, refresh = 0, save_warmup = FALSE
)
```

Posterior estimates of the fixed effects:
```{r}
plot_betas(gaussian_example_fit)
```

Posterior estimates of time-varying effects
```{r}
plot_deltas(gaussian_example_fit, scales = "free")
```

And group-specific intercepts:
```{r}
plot_nus(gaussian_example_fit)
```

## Related packages
The `dynamite` package uses Stan via `rstan` (see https://mc-stan.org).
