---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "80%"
)
```

```{r srr-tags, eval = FALSE, echo = FALSE}
#' @srrstats {G1.1,  G1.2}
```

# dynamite

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/santikka/dynamite/workflows/R-CMD-check/badge.svg)](https://github.com/santikka/dynamite/actions)
[![Codecov test coverage](https://codecov.io/gh/santikka/dynamite/branch/main/graph/badge.svg)](https://app.codecov.io/gh/santikka/dynamite?branch=main)
<!-- badges: end -->

The `dynamite` [R](https://www.r-project.org/) package provides easy-to-use interface for Bayesian inference of complex panel (time series) data comprising of multiple measurements per multiple individuals measured in time. The main features distinguishing the package and the underlying methodology from many other approaches are:

* Support for both time-invariant and time-varying effects modeled via 
  B-splines.
* Joint modeling of multiple measurements per individual (multiple channels) 
  based directly on the assumed data generating process.
* Support for non-Gaussian observations: Currently Gaussian, Categorical, 
  Poisson, Bernoulli, Binomial, Negative Binomial, Gamma, Exponential, and Beta 
  distributions are available and these can be mixed arbitrarily in 
  multichannel models.
* Allows evaluating realistic long-term counterfactual predictions which take 
  into account the dynamic structure of the model by posterior predictive 
  distribution simulation.
* Transparent quantification of parameter and predictive uncertainty due to a 
  fully Bayesian approach.
* User-friendly and efficient R interface with state-of-the-art estimation via 
  Stan.

The `dynamite` package is developed with the support of Academy of Finland 
grant 331817 ([PREDLIFE](https://sites.utu.fi/predlife/en/)). 

## Installation

`dynamite` uses R's native pipe operator, so R version 4.1.0 or newer is required.

`dynamite` also relies on some of the features of [Stan](https://mc-stan.org/)'s 
which are not yet available on the [CRAN](https://cran.r-project.org/) version 
of [rstan](https://CRAN.R-project.org/package=rstan) pacakge. Therefore you 
need to install latest `rstan` and `StanHeaders` packages as
```{r, eval = FALSE}
# run the next line if you already have rstan installed:
# remove.packages(c("StanHeaders", "rstan"))
install.packages("rstan", repos = c("https://mc-stan.org/r-packages/", 
  getOption("repos")))
```
(See more help on [RStan wiki](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started))

Finally, a version `1.14.3` or newer of the 
[data.table](https://github.com/Rdatatable/data.table) package is required, 
which can be installed as

```{r, eval = FALSE}
# run the next line if you do not have any version of the data.table installed:
# install.packages("data.table")

# install the latest development version:
data.table::update.dev.pkg()
```

After these steps, you can install the development version of dynamite from 
[GitHub](https://github.com/) with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("santikka/dynamite")
```

## Example

A single-channel model with time-invariant effect of `z`, time-varying effect 
of `x`, lagged value of the response variable `y` and a group-specific random 
intercepts:

```{r, echo = FALSE}
library(dynamite)
```

```{r, eval = FALSE}
set.seed(1)
library(dynamite)
gaussian_example_fit <- dynamite(
  obs(y ~ -1 + z + varying(~ x + lag(y)), family = "gaussian") + 
    random() + splines(df = 20),
  data = gaussian_example, time = "time", group = "id",
  iter = 2000, warmup = 1000, thin = 5,
  chains = 2, cores = 2, refresh = 0, save_warmup = FALSE
)
```

Posterior estimates of the fixed effects:
```{r}
plot_betas(gaussian_example_fit)
```

Posterior estimates of time-varying effects
```{r}
plot_deltas(gaussian_example_fit, scales = "free")
```

And group-specific intercepts:
```{r}
plot_nus(gaussian_example_fit)
```

Traceplots and density plots:
```{r}
plot(gaussian_example_fit, type = "beta")
```

Posterior predictive samples for first 4 groups (samples based on the posterior
distribution of model parameters and observed data on first time point):
```{r}
library(ggplot2)
pred <- predict(gaussian_example_fit, n_draws = 50)
pred |> dplyr::filter(id < 5) |> 
  ggplot(aes(time, y_new, group = draw)) +
  geom_line(alpha = 0.5) + 
  # observed values
  geom_line(aes(y = y), colour = "tomato") +
  facet_wrap(~ id) +
  theme_bw()
```

For more examples, see the package vignette.

## Related packages

- The `dynamite` package uses Stan via [`rstan`](https://CRAN.R-project.org/package=rstan) (see also https://mc-stan.org), which is a probabilistic programming language for general Bayesian modelling.

- The [`brms`](https://CRAN.R-project.org/package=brms) package also uses Stan, and can be used to fit various complex multilevel models.

- Regression modelling with time-varying coefficients based on kernel smoothing and least squares estimation is available in package [`tvReg`](https://CRAN.R-project.org/package=tvReg). The [`tvem`](https://CRAN.R-project.org/package=tvem) package provides similar functionality for gaussian, binomial and poisson responses with [`mgcv`](https://CRAN.R-project.org/package=mgcv) backend.

- [`plm`](https://CRAN.R-project.org/package=plm) contains various methods to estimate linear models for panel data, e.g. the fixed effect models.

- [`lavaan`](https://CRAN.R-project.org/package=lavaan) provides tools for structural equation modelling, and as such can be used to model various panel data models as well.


